Symmetric Peer-to-Peer Applications

### 2022-10-29 final/p2p-22.tex
...

\usepackage{xspace}
\newcommand{\FC}       {Freechains\xspace}
\newcommand{\reps}     {\emph{reps}\xspace}
\newcommand{\onerep}   {\emph{1~rep}\xspace}
\newcommand{\nreps}[1] {\emph{#1~reps\xspace}}
\newcommand{\code}[1]  {\texttt{\footnotesize{#1}}}
\newcommand{\Xon} {$1{\rightarrow}N$\xspace}
\newcommand{\Xno} {$1{\leftarrow}N$\xspace}
\newcommand{\Xnn} {$N{\leftrightarrow}N$\xspace}
\newcommand{\Xoo} {$1{\leftrightarrow}1$\xspace}
\newcommand{\Xo}  {$1{\hookleftarrow}$\xspace}
\newcommand{\lin}[1]{(\emph{ln. #1}\xspace)}
\newcommand{\linx}[1]{\emph{ln. #1}\xspace}

@@

\abstract{
In real-time networked applications, such as shared documents, users can
interact remotely and yet share the exact same experience as if they were in a
single machine.
%
In this work, we propose a middleware for \emph{symmetric peer-to-peer
applications} in which decentralized instances can broadcast asynchronous
events and yet conform to identical behavior.
%
Peers are allowed to join and leave the network at any time.
Application developers must adhere to a restricted API, which is deterministic,
stateless, and only supports pre-allocated memory.
%
The middleware is responsible for synchronizing the events in a global shared
timeline across the network.
Also, based on memory snapshots and deterministic simulation, a ``time machine''
can rollback conflicting peers to resynchronize their state.
%
We show that the middleware can handle applications with over 20 peers in a
heterogeneous topology under high churn.
For instance, in a simulation scenario of 25 events per minute with a network
latency of $50ms$, the peers need to roll back and resynchronize only $3\%$ of
the time.
}

@@

Real-time networked applications allow multiple users to interact remotely and
yet share the same experience.
Examples of these \emph{symmetric distributed applications}~\cite{gals} are
shared documents, watch parties, and multi-player games.
%
In order to reproduce the exact behavior in multiple devices, the application
must be able to synchronize time and execution across the network.
In addition, since users can interact asynchronously with the application,
event occurrences must somehow be synchronized back across devices.

A common approach towards symmetric applications is to introduce a middleware
to orchestrate events and time in the network~\cite{gals,croquet}.
This way, applications developers can rely on middleware primitives to trigger
events, which are intercepted and synchronized in a global shared timeline
across the network.
Developers must also restrict themselves to deterministic and stateless calls
only, such that execution can be equally reproduced in all nodes according to
the shared timeline.
However, current solutions depend on a central server to interconnect network
nodes and determine a shared timeline.

In this work, we propose \emph{symmetric peer-to-peer applications}, which
do not rely on central servers for coordination.
Peers in the application form a dynamic network graph and communicate only
with direct neighbours, as in typical unstructured peer-to-peer networks.
Events are flooded in the graph and are triggered locally with a small delay
to compensate the network latency.
To deal with events received out of order or too late, a distributed time
machine can rollback peers to a previous state and reapply events in order and
in time.
Our main contribution is the design of a middleware to ensure that all peers
    (a) meet event deadlines,
    (b) advance in sync in real time, and
    (c) can leave and join the network and remain symmetric.

We perform simulations with over 20 peers in a 5-hop mixed topology with cycles
and straight lines.
We vary the network latency and peer churn, as well as the frequency and
deadlines of events.
%
We then measure (a) the recurrence of time travels, (b) the real-time pace of
peers, and (c) the recoverability from churns, such that they validate our
goals above, respectively.
%
As results, we show that the last two goals are met even under extreme
conditions: peers exhibit a time mismatch of under $0.2\%$ and take less than
$1.5s$ to recover from churns.
%
For the first goal, we identify the scenarios in which the middleware can meet
event deadlines with the desired performance.
As an example, in a scenario of 25 events per minute with a network latency of
$50ms$, the peers need to roll back only $3\%$ of the time.

@@

% docs.google.com/spreadsheets/d/1CpMmEgabJq2XQeTDW_BNCV4JWrcMlELuaEjDtw7kwg0/
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{table}
  \caption{
    Related work regarding
        (a) how the network is organized,
        (b) how global time is determined, and
        (c) how events are propagated and applied.
    \label{fig.table}
  }
\end{figure}

@@

Croquet%
\footnote{Croquet.io: https://croquet.io/}~\cite{croquet} guarantees
bit-identical real-time behavior for users in collaborative distributed
environments.
%
As detailed in Figure~\ref{fig.algos}, a centralized \emph{reflector} issues
periodic ticks, such that all nodes in the network advance together according
to a synchronized global shared clock.
If a user generates an event, it is sent back to the reflector, which
broadcasts the event in the next tick, which all nodes apply in sync.
%
Croquet takes periodic snapshots of the whole application state in order to
support late joins to a running session.
The new node just needs to restore the latest snapshot and simulate the
remaining events to reach the current running state.
%
As indicated in Figure~\ref{fig.table}, Croquet relies on a centralized
network (a), in which nodes advance in sync (b), and in which event outcomes
depend on the central server to be online (c).

GALS~\cite{gals} shares the same goals with Croquet, but with some tradeoffs,
mostly favoring clients with slow connections.
As detailed in Figure~\ref{fig.algos}, instead of advancing ticks in sync with
the server, clients have their own independent local clocks.
Event generation requires two round trips (\emph{when→ask→ack→tick}) and is
delayed dynamically according to the slowest client.
%
On the one hand, ticks do not generate any traffic and clients have smooth
frame transitions, even those with poor connections.
On the other hand, events take longer to apply and clients may experience
occasional freezes.
The syncing protocol also requires extra bookkeeping to deal with clock drifts
and client disconnections~\cite{gals}.
%
As indicated in Figure~\ref{fig.table}, GALS also relies on a centralized
network (a), but in which nodes advance time independently (b), and in which
event outcomes still depend on the central server to be online (c).

@@

An antagonistic approach to deterministic reactions to events is to model the
application with conflict-free replicated data types~\cite{crdts}.
CRDTs are designed in such a way that all operations are commutative, so that
the order in which they are applied does not affect the final outcome.
%
As detailed in Figure~\ref{fig.algos}, peers can communicate operations
directly to each other with no central authority.
Also, since operations need not to be ordered, they can be applied at the
very first moment they are generated or received, even if the peers are
offline.
%
Since operations must be commutative, it is not possible to timestamp them
in a unique global shared clock.
Hence, there is no notion of a timeline in which peers go through
bit-identical steps, but they do eventually reach the same state.
%
On the one hand, the main advantage of CRDTs is to support local-first
software~\cite{local}, since they can work offline in the same way as online.
On the other hand, they provide very restrictive APIs (the CRDTs themselves),
and do not support real-time consensus among peers.
%
As indicated in Figure~\ref{fig.table}, CRDTs work on decentralized networks
(a), in which peers advance independently (b), and in which operations can be
applied immediately, even while offline (c).
%
Automerge along with its accompanying Hypermerge peer-to-peer infrastructure
is an example of decentralized CRDTs~\cite{p2p.automerge,p2p.pushpin}.

In this work, as indicated in Figure~\ref{fig.table}, our goal is to support
peer-to-peer real-time symmetric execution (a), while still tolerating
out-of-order (b) and offline (c) event generation.
%
As detailed in Figure~\ref{fig.algos}, peers have independent timelines and
can communicate events directly to each other.
Events are delayed to be able to reach other peers in time.
In the case a deadline is missed, the peer rolls back in time and then fast
forwards execution until it catches real-time behavior again.
%
Like Croquet and GALS, peers compute events in sequence using deterministic and
stateless calls only.

We now discuss related work on time machines.
%
Regarding decentralized machines, Fusion%
\footnote{Fusion: \url{https://doc.photonengine.com/en-us/fusion}}
is a state synchronization networking library for the Unity game engine.
It provides networked tick-based simulation in two modes: client-server
(\emph{hosted mode}) and peer-to-peer (\emph{shared mode}).
The hosted mode is based on continuous memory snapshots, which allows for
full rollbacks when clients diverge.
However, the shared mode is less powerful and only supports eventual
consistency among clients.
We presume that continuous memory snapshots would be too costly without a
central server.
%
Regarding single-user local machines, the video game Braid%
\footnote{Braid: \url{http://braid-game.com/}}
is designed on the assumption that players have unrestricted rollbacks as part
of the game mechanics.
Like Fusion, the implementation is based on continuous memory snapshots,
instead of deterministic simulation as we propose in this work.
%
Schuster~\cite{tml.js} proposes a simple API for time travelling in
event-driven JavaScript:
    \code{init()} to return an initial state;
    \code{handle(evt,mem)} to process events and modify state; and
    \code{render(mem)} to output the current state.
As we describe next, we use a similar approach by separating state modification
from rendering~\cite{tml.alive}.

@@

We guide our discussion through the didactic example of Figure~\ref{fig.move},
in which multiple users use the arrow keys to control the direction of a
moving rectangle.
The six instances are connected in a peer-to-peer network, such that most
peers can only communicate indirectly to each other.
%
The application is symmetric in the sense that if any user presses a key, a
corresponding event is broadcast and applied in all peers simultaneously, as
if they were a single mirrored machine.
At any moment, if a user pauses the application, all peers are guaranteed to
pause exactly at the same frame with the rectangle being ``bit-identically'' at
the same position.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{middleware}
  \caption{
    Applications generate asynchronous events and communicate through an API
    with the middleware.
    The middleware orchestrates the peers, synchronizing their ticks locally
    and broadcasting events with a $\Delta$ delay.
    \label{fig.middleware}
  }
\end{figure}

In Figure~\ref{fig.middleware}, a peer is depicted as an application that uses
an API to communicate with the middleware.
The application itself is not aware of the network and has no direct access to
other peers, which communicate transparently through the middleware.
The application generates asynchronous events that go through the middleware,
which synchronizes them with a timestamp with a $\Delta$ delay scheduled to the
future, such that all peers can satisfy.
The middleware controls the execution of the application by issuing ticks and
synchronized events.

The main job of the middleware is to deal with the uncertainties of the
network in order to preserve the symmetric behavior across peers.
As described in the Introduction, the middleware ensures that all peers
    (a) meet event deadlines,
    (b) advance in sync in real time, and
    (c) can leave and join the network and remain symmetric:
%
\begin{enumerate}
\item \textbf{Event Deadlines:}
Due to the inherent latency of networks, peers generate events with a
deliberate $\Delta$ so that they can reach other peers in time to be applied in
sync.
%We first assume a predetermined delay for the sake of simplicity.
Nevertheless, a peer ahead of time may receive an event that should have been
applied in the past, even considering the delay.
In this case, the middleware automatically rolls back the peer, applies the
event at the correct time, and then fast forwards the peer, re-applying all
remaining events up to the real time.
As we detail further, rolling back a peer requires to simulate the whole
application execution from the beginning.
As an optimization, the middleware takes periodic snapshots locally to avoid
full simulation.
%
\item \textbf{Time Synchronization:}
Since peers run in different machines, their timelines will inevitably be out
of sync because applications are launched locally at different times, and
because internal clocks may diverge over time (e.g., from rollbacks or timer
polling inaccuracies).
The middleware assumes the maximum time among all peers the be the
``correct real time'', so that other peers must fast forward to catch up with
it.
Therefore, in order to determine the maximum time and synchronize the clocks,
the middleware proceeds as follows:
    (a) peers broadcast \code{SYNC} events periodically with their local
        times, and
    (b) if a peer is behind a received \code{SYNC}, it advances its frames
        proportionally faster to catch up in time.
%
\item \textbf{Peer Churn:}
Considering that we target peer-to-peer networks, it is important that the
middleware can handle arrival and departure of peers seamlessly.
Note that high churn may lead to intermittent network partitions.
Given our unstructured approach, nothing needs to be done when a peer leaves
the network.
However, when a peer joins the network, the middleware needs to ensure that it
receives from other peer all events that ever happened, in order.
This is trivial since all peers already need to hold the full event history (as
described in item (a)), including their absolute timestamps in the shared
timeline.
\end{enumerate}

@@

{\footnotesize
\begin{verbatim}
int main (void) {
    p2p_loop (
        1,              // peer identifier
        50,             // simulation FPS
        sizeof(G), &G,  // pre-allocated full state
        cb_ini,         // init/quit callback
        cb_sim,         // simulation callback
        cb_ren,         // rendering callback
        cb_evt          // events callback
    );
}
\end{verbatim}
}

The middleware assumes that each peer assigns itself a unique identifier in a
contiguous range.
The FPS must be the same in all peers to ensure bit-identical simulation.

@@

The variable \code{G} passed to the middleware holds the full application
state.
This allows the middleware to take memory snapshots and rollback to previous
states.
If dynamic allocation is required, it is necessary to hold a finite memory
pool in \code{G} with a custom allocator.
In our guiding example of Figure~\ref{fig.move}, the moving rectangle
application only requires to hold the current position and direction:

{\footnotesize
\begin{verbatim}
struct {
    int x,  y;   // position
    int dx, dy;  // direction speed
} G;
\end{verbatim}
}

@@

{\footnotesize
\begin{verbatim}
enum {
    P2P_EVT_KEY = P2P_EVT_NEXT // next to predefined
};
\end{verbatim}
}

@@

{\footnotesize
\begin{verbatim}
void cb_ini (int ini) {
    if (ini) {
        <...> // create the SDL window
        <...> // open the arrow PNG images
        // create peer links (peer-id, ip, port)
        p2p_link(2, "192.168.1.2", 5000);
    } else {
        <...> // destroy the SDL window
        <...> // destroy the arrow PNG images
        // destroy peer links
        p2p_unlink(2);
    }
}
\end{verbatim}
}

@@

The middleware calls \code{cb\_sim} every frame or event occurrence, which
should affect the simulation state \code{G} deterministically.
An occurring event may be passed as argument to the callback, which must never
perform any side effects, such as stateful calls or rendering frames.
In our example, we modify the state of the rectangle as follows:
    (a) reset its position and speed on \code{START},
    (b) increment its position based on the current speed on \code{TICK}, and
    (c) modify its speed based on \code{KEY}:

{\footnotesize
\begin{verbatim}
void cb_sim (p2p_evt evt) {
    switch (evt.id) {
        case P2P_EVT_START: // resets pos and speed
            G.x = G.y = G.dx = G.dy = 0;
            break;
        case P2P_EVT_TICK:  // increases position
            G.x += G.dx * VEL;
            G.y += G.dy * VEL;
            break;
        case P2P_EVT_KEY:   // modifies speed
            switch (evt.pay.i1) {
                case SDLK_UP:
                    G.dx = 0;
                    G.dy = -1;
                    break;
                <...> // same for other keys
            }
            break;
    }
}
\end{verbatim}
}

@@

Time travelling is also the reason why \code{cb\_sim} must be separated from
the rendering callback \code{cb\_ren}~\cite{tml.js}, which would otherwise
render the screen multiple times while travelling.
%, which the middleware calls only on real-time frames.
In our example, \code{cb\_ren} just needs to redraw the rectangle at the
current position:

{\footnotesize
\begin{verbatim}
void cb_ren (void) {
    <...> // clears the SDL window
    // redraws the rectangle at the current position
    SDL_Rect r = { G.x, G.y, 20, 20 };
    SDL_DrawRect(&r);
}
\end{verbatim}
}

@@

{\footnotesize
\begin{verbatim}
int cb_evt (SDL_Event* sdl, p2p_evt* evt) {
    if (sdl->type == SDL_KEYDOWN) {
        if (sdl->key == <any-of-the-keys>) {
            SDL_DrawImage(<img-of-the-key>);
            *evt = (p2p_evt){ P2P_EVT_KEY,sdl->key };
            return 1; // generate application event
        }
    }
    return 0; // otherwise, do not generate any event
}
\end{verbatim}
}

@@

As illustrated in Figure~\ref{fig.middleware}, events are broadcast between
peers with a timestamp scheduled to the future with a $\Delta$ such that all
peers are able to apply them in sync.
%
To prevent broadcast cycles, each peer keeps a vector with the highest
timestamp it received from each other peer, such that lower numbers can be
ignored when received.
Since we rely on TCP connections, it is not possible to receive out-of-order
timestamps attached to a given source peer.
When the middleware receives a higher number from a given peer, it updates its
vector position and triggers a broadcast to neighbours.
%
As mentioned in Section~\ref{sec.tml.api}, we assume that peers assign
themselves unique identifiers in a contiguous range, such that they fit in a
simple vector of timestamps.
Therefore, this is not a severe limitation, and a separate mechanism could be
used for assignments, with few modifications to the core of the middleware.

@@

{\footnotesize
\begin{verbatim}
typedef struct {
    uint8_t  src;       // source peer
    uint32_t tick;      // tick to apply
    p2p_evt  evt;       // event id + payload
} p2p_pak;

struct {
    int i;  // next event to apply in real time
    int n;  // number of events in the queue
    p2p_pak buf[MAX];   // queue of packets
} PAKS;
\end{verbatim}
}

@@

{\footnotesize
\begin{verbatim}
01  void p2p_travel (int from, int to, int ms) {
02    for (i=from..to) {
03      int bef = <tick-of-snapshot-before-i>
04      <restore-snapshot-at-bef>
05      for (j=bef..i) {
06        cb_sim(<tick-or-evt-at-j>)
07      }
08      cb_ren();
09      <delay-ms>
10    }
11  }
\end{verbatim}
}

@@

This section provided an overview of the middleware implementation to answer
how it ensures the desired properties in concrete terms as follows:
    (a) peers meet event deadlines either travelling back and forth or in real
        time (\linx{12--16} or \lin{18--22} / \ref{sec.tml.middleware.loop});
    (b) peers synchronize their clocks with periodic \code{SYNC} packets that
        fast forward late peers
        (\linx{7--10} / \ref{sec.tml.middleware.loop}); and
    (c) peers can join the network at any time and still catch up in time by
        receiving the full event history
        (Section~\ref{sec.tml.middleware.events}).

>>>>

\usepackage{xspace}
\newcommand{\FC}       {Freechains\xspace}
\newcommand{\reps}     {\emph{reps}\xspace}
\newcommand{\onerep}   {\emph{1~rep}\xspace}
\newcommand{\nreps}[1] {\emph{#1~reps\xspace}}
\newcommand{\code}[1]  {\texttt{\small{#1}}}
\newcommand{\Xon} {$1{\rightarrow}N$\xspace}
\newcommand{\Xno} {$1{\leftarrow}N$\xspace}
\newcommand{\Xnn} {$N{\leftrightarrow}N$\xspace}
\newcommand{\Xoo} {$1{\leftrightarrow}1$\xspace}
\newcommand{\Xo}  {$1{\hookleftarrow}$\xspace}
\newcommand{\lin}[1]{(\emph{ln. #1}\xspace)}
\newcommand{\linx}[1]{\emph{ln. #1}\xspace}

@@

\abstract{
In real-time networked applications, such as collaborative documents, users can
interact remotely and yet share the exact same experience as if they were in a
single machine.
%
In this work, we propose a middleware for \emph{symmetric peer-to-peer
applications} in which decentralized instances can broadcast asynchronous
events and yet conform to identical behavior.
%
Peers are allowed to join and leave the network at any time.
Application developers must adhere to a restricted API, which is deterministic,
stateless, and only supports pre-allocated memory.
%
The middleware is responsible for synchronizing the events in a global shared
timeline across the network.
Also, based on memory snapshots and deterministic simulation, a ``time machine''
can rollback conflicting peers to resynchronize their states.
%
We show that the middleware can handle applications with over 20 peers in a
heterogeneous topologies under high churns.
For instance, in a simulation scenario with a network latency of 50ms and 25
events per minute, the peers need to roll back and resynchronize only 3\% of
the time.
}

@@

Real-time networked applications allow multiple users to interact remotely and
yet share the same experience.
Examples of these \emph{symmetric distributed applications}~\cite{gals} are
collaborative documents, watch parties, and multi-player games~\cite{TODO:collab}.
%
In order to reproduce the exact behavior in multiple devices, the application
must be able to synchronize time and execution across the network.
In addition, since users can interact asynchronously with the application,
event occurrences must somehow be synchronized back across devices.

A common approach towards symmetric applications is to introduce a middleware
to orchestrate events and time in the network~\cite{gals,croquet}.
This way, applications developers can rely on middleware primitives to trigger
events, which are intercepted and synchronized in a global shared timeline
across the network.
Developers must also restrict themselves to deterministic and stateless calls
only, such that execution can be equally reproduced in all nodes according to
the shared timeline~\cite{TODO:det_reprod}.
However, current solutions depend on a central server to interconnect network
nodes and determine a shared timeline.

In this work, we propose \emph{symmetric peer-to-peer applications}, which
do not rely on central servers for coordination.
Peers in the application form a dynamic network graph and communicate only
with direct neighbours, as in typical unstructured peer-to-peer
networks~\cite{TODO:p2p}.
Events are flooded in the graph and are triggered with a small deadline delay
to compensate the network latency.
To deal with events received out of order or too late, a distributed time
machine can rollback peers to a previous state and reapply events in order and
in time.
Our main contribution is the design of a middleware to ensure that all peers
    (a) meet event deadlines,
    (b) advance in sync in real time, and
    (c) can leave and join the network and remain symmetric.

We perform simulations with over 20 peers in a 5-hop mixed topology with cycles
and straight lines.
We vary the network latency and peer churn, as well as the frequency and
deadlines of events.
%
We then measure (a) the recurrence of time travels, (b) the real-time pace of
peers, and (c) the recoverability from churns, such that they validate our
goals above, respectively.
%
As results, we show that the last two goals are met even under extreme
conditions: peers exhibit a time mismatch of under $0.2\%$ and take less than
$1.5s$ to recover from churns.
%
For the first goal, we identify the scenarios in which the middleware can meet
event deadlines with the desired performance.
As an example, in a scenario with a network latency of $50ms$ and 25 events per
minute, the peers need to roll back only $3\%$ of the time.

@@

% docs.google.com/spreadsheets/d/1CpMmEgabJq2XQeTDW_BNCV4JWrcMlELuaEjDtw7kwg0/
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{table}
  \caption{
    Related work regarding
        (a) how the network is organized,
        (b) how global time is determined, and
        (c) how events are broadcast and applied.
    \label{fig.table}
  }
\end{figure}

@@

Croquet%
\footnote{Croquet.io: https://croquet.io/}~\cite{croquet} guarantees
bit-identical real-time behavior for users in collaborative distributed
environments.
%
As detailed in Figure~\ref{fig.algos}, a centralized \emph{reflector} issues
periodic ticks, such that all nodes in the network advance together according
to a synchronized global shared clock.
If a user generates an event, it is sent back to the reflector, which
broadcasts the event in the next tick, which all nodes apply in sync.
%
Croquet takes periodic snapshots of the whole application state in order to
support late joins to a running session.
The new node just needs to restore the latest snapshot and simulate the
remaining events to reach the current running state.
%
As indicated in Figure~\ref{fig.table}, Croquet relies on a centralized
network (a), in which nodes advance in sync (b), and in which event broadcasts
and outcomes depend on the central server to be online (c).

GALS~\cite{gals} shares the same goals with Croquet, but with some tradeoffs,
mostly favoring clients with slow connections.
As detailed in Figure~\ref{fig.algos}, instead of advancing ticks in sync with
the server, clients have their own independent local clocks.
Event generation requires two round trips (\emph{when→ask→ack→tick}) and is
delayed dynamically according to the slowest client.
%
On the one hand, ticks do not generate any traffic and clients have smooth
frame transitions, even those with poor connections.
On the other hand, events take longer to apply and clients may experience
occasional freezes.
The syncing protocol also requires extra bookkeeping to deal with clock drifts
and client disconnections~\cite{gals}.
%
As indicated in Figure~\ref{fig.table}, GALS also relies on a centralized
network (a), but in which nodes advance time independently (b), and in which
event broadcasts and outcomes still depend on the central server to be online
(c).

@@

An antagonistic approach to deterministic reactions to events is to model the
application with conflict-free replicated data types~\cite{crdts}.
CRDTs are designed in such a way that all operations are commutative, so that
the order in which they are applied does not affect the final outcome.
%
As detailed in Figure~\ref{fig.algos}, peers can communicate operations
directly to each other with no central authority.
Also, since operations need not to be ordered, they can be applied at the
very first moment they are generated or received, even if the peers are
offline.
%
Since operations must be commutative, it is not possible to timestamp them
in a unique global shared clock.
Hence, there is no notion of a timeline in which peers go through
bit-identical steps, but they do eventually reach the same
state~\cite{TODO:eventual}.
%
On the one hand, the main advantage of CRDTs is to support local-first
software~\cite{local}, since they can work offline in the same way as online.
On the other hand, they provide very restrictive APIs (the CRDTs themselves),
and do not support real-time consensus among peers~\cite{TODO:consensus}.
%
As indicated in Figure~\ref{fig.table}, CRDTs work on decentralized networks
(a), in which peers advance independently (b), and in which operations can be
applied immediately, even while offline (c).
%
Automerge along with its accompanying Hypermerge peer-to-peer infrastructure
is an example of decentralized CRDTs~\cite{p2p.automerge,p2p.pushpin}.

In this work, as indicated in Figure~\ref{fig.table}, our goal is to support
peer-to-peer (a) real-time symmetric execution, while still tolerating
out-of-order (b) and offline (c) event generation.
%
As detailed in Figure~\ref{fig.algos}, peers have independent timelines and
can communicate events directly to each other.
Event deadlines are delayed to be able to reach other peers in time.
In the case a deadline is missed, the peer rolls back in time and then fast
forwards execution until it catches real-time behavior again.
%
Like Croquet and GALS, peers compute events in sequence using deterministic and
stateless calls only.

We now discuss related work on time machines.
%
Regarding decentralized machines, Fusion~\cite{fusion} is a state
synchronization networking library for the Unity game engine.
It provides networked tick-based simulation in two modes: client-server
(\emph{hosted mode}) and peer-to-peer (\emph{shared mode}).
The hosted mode is based on continuous memory snapshots, which allows for
full rollbacks when clients diverge.
However, the shared mode is less powerful and only supports eventual
consistency among clients.
We presume that continuous memory snapshots would be too costly without a
central server.
%
Regarding single-user local machines, the video game Braid%
\footnote{Braid: \url{http://braid-game.com/}}
is designed on the assumption that players have unrestricted rollbacks as part
of the game mechanics.
Like Fusion, the implementation is based on continuous memory snapshots,
instead of deterministic simulation as we propose in this work.
%
Schuster~\cite{tml.js} proposes a simple API for time travelling in
event-driven JavaScript:
    \code{init()} to return an initial state;
    \code{handle(evt,mem)} to process events and modify state; and
    \code{render(mem)} to output the current state.
As we describe next, we use a similar approach by separating state mutation
from rendering~\cite{tml.alive}.

@@

We guide our discussion through the didactic example of Figure~\ref{fig.move},
in which multiple users use the arrow keys to control the direction of a
moving rectangle.
The six instances are connected in a peer-to-peer network, such that most
peers can only communicate indirectly to each other.
%
The application is symmetric in the sense that if any user presses a key, a
corresponding event is broadcast and applied in all peers simultaneously, as
if they were a single mirrored machine.
At any moment, if a user pauses the application, all peers are guaranteed to
pause exactly at the same frame, with the rectangle ``bit-identically'' at the
same position.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{middleware}
  \caption{
    Applications generate asynchronous events and communicate through an API
    with the middleware.
    The middleware orchestrates the peers, synchronizing their ticks locally
    and broadcasting events with a deadline delay ($\Delta$).
    \label{fig.middleware}
  }
\end{figure}

In Figure~\ref{fig.middleware}, a peer is depicted as an application that uses
an API to communicate with the middleware.
The application itself is not aware of the network and has no direct access to
other peers, which communicate transparently through the middleware.
The application generates asynchronous events that go through the middleware,
which synchronizes them with a timestamp with a deadline delay ($\Delta$)
scheduled to the future, such that all peers can satisfy.
The middleware controls the execution of the application by issuing ticks and
synchronized events.

The main job of the middleware is to deal with the uncertainties of the
network in order to preserve the symmetric behavior across peers.
As described in the Introduction, the middleware ensures that all peers
    (a) meet event deadlines,
    (b) advance in sync in real time, and
    (c) can leave and join the network and remain symmetric:
%
\begin{enumerate}
\item \textbf{Event Deadlines:}
Due to the inherent latency of networks, peers generate events with a
deliberate $\Delta$ so that they can reach other peers in time to be applied in
sync.
%We first assume a predetermined delay for the sake of simplicity.
Nevertheless, a peer ahead of time may receive an event that should have been
applied in the past, even considering its $\Delta$.
In this case, the middleware automatically rolls back the peer, applies the
event at the correct time, and then fast forwards the peer, re-applying all
remaining events up to the real time.
As we detail further, rolling back a peer requires to simulate the whole
application execution and event occurrences from the beginning.
As an optimization, the middleware takes periodic snapshots locally to avoid
full simulation.
%
\item \textbf{Time Synchronization:}
Since peers run in different machines, their timelines will inevitably be out
of sync because applications are launched locally at different times, and
because internal clocks may diverge over time (e.g., from rollbacks or timer
polling inaccuracies).
The middleware assumes the maximum time among all peers the be the
``correct real time'', so that other peers must fast forward to catch up with
it.
Therefore, in order to determine the maximum time and synchronize the clocks,
the middleware proceeds as follows:
    (a) peers broadcast \code{SYNC} events periodically with their local
        times, and
    (b) if a peer is behind a received \code{SYNC}, it advances its frames
        proportionally faster to catch up in time.
%
To avoid abrupt visual changes, the middleware makes smooth time transitions
that take up to $1s$ to complete.
%
\item \textbf{Peer Churn:}
Considering that we target dynamic peer-to-peer networks, it is important that
the middleware can handle arrival and departure of peers seamlessly.
Note that high churn may lead to intermittent network partitions.
Given our unstructured approach, nothing needs to be done when a peer leaves
the network.
However, when a peer joins the network, the middleware needs to ensure that it
receives from other peer all events that ever happened, in order.
This is trivial since all peers already need to hold the full event history (as
described in item (a)).
% including their absolute timestamps in the shared timeline.
\end{enumerate}

@@

{\footnotesize
~
\begin{verbatim}
int main (void) {
    p2p_loop (
        1,              // peer identifier
        50,             // simulation FPS
        sizeof(G), &G,  // pre-allocated full state
        cb_ini,         // init/quit callback
        cb_sim,         // simulation callback
        cb_ren,         // rendering callback
        cb_evt          // events callback
    );
}
\end{verbatim}
~
}

The middleware assumes that each peer assigns itself a unique identifier, such
as TODO~\cite{TODO:hash}.
The FPS must be the same in all peers to ensure bit-identical simulation.

@@

The variable \code{G} passed to the middleware holds the full application
state.
This allows the middleware to take memory snapshots and rollback to previous
states.
If dynamic allocation is required, it is necessary to hold a finite memory
pool in \code{G} with a custom allocator.
In our guiding example of Figure~\ref{fig.move}, the application only needs to
hold the rectangle position and direction:

{\footnotesize
~
\begin{verbatim}
struct {
    int x,  y;   // position
    int dx, dy;  // direction speed
} G;
\end{verbatim}
~
}

@@

{\footnotesize
~
\begin{verbatim}
enum {
    P2P_EVT_KEY = P2P_EVT_NEXT // key is next to predefs
};
\end{verbatim}
~
}

@@

{\footnotesize
~
\begin{verbatim}
void cb_ini (int ini) {
    if (ini) {
        <...> // create the SDL window
        <...> // open the arrow PNG images
        // create peer links (peer-id, ip, port)
        p2p_link(2, "192.168.1.2", 5000);
    } else {
        <...> // destroy the SDL window
        <...> // destroy the arrow PNG images
        // destroy peer links
        p2p_unlink(2);
    }
}
\end{verbatim}
~
}

@@

The middleware calls \code{cb\_sim} every frame or event occurrence, which
should affect the simulation state \code{G} deterministically.
An occurring event may be passed as argument to the callback, which must never
perform any side effects, such as stateful calls or rendering frames.
In our example, we modify the state of the rectangle as follows:
    on \code{START}, reset its position and speed;
    on \code{TICK},  increment its position based on the current speed; and
    on \code{KEY},   modify its speed:

{\footnotesize
~
\begin{verbatim}
void cb_sim (p2p_evt evt) {
    switch (evt.id) {
        case P2P_EVT_START: // resets pos and speed
            G.x = G.y = G.dx = G.dy = 0;
            break;
        case P2P_EVT_TICK:  // increases position
            G.x += G.dx * VEL;
            G.y += G.dy * VEL;
            break;
        case P2P_EVT_KEY:   // modifies speed
            switch (evt.pay.i1) {
                case SDLK_UP:
                    G.dx = 0;
                    G.dy = -1;
                    break;
                <...> // similar for other keys
            }
            break;
    }
}
\end{verbatim}
~
}

@@

Time travelling is also the reason why \code{cb\_sim} must be separated from
the rendering callback \code{cb\_ren}~\cite{tml.js}, which would otherwise
render the screen multiple times while travelling.
The middleware renders the screen only after each frame in real time.
%, which the middleware calls only on real-time frames.
In our example, \code{cb\_ren} just needs to redraw the rectangle at the
current position:

{\footnotesize
~
\begin{verbatim}
void cb_ren (void) {
    <...> // clears the SDL window
    // redraws the rectangle at the current position
    SDL_Rect r = { G.x, G.y, 20, 20 };
    SDL_DrawRect(&r);
}
\end{verbatim}
~
}

@@

{\footnotesize
~
\begin{verbatim}
int cb_evt (SDL_Event* sdl, p2p_evt* evt) {
    if (sdl->type == SDL_KEYDOWN) {
        if (sdl->key == <any-of-the-keys>) {
            SDL_DrawImage(<img-of-the-key>);
            *evt = (p2p_evt){ P2P_EVT_KEY,sdl->key };
            return 1; // generate application event
        }
    }
    return 0; // otherwise, do not generate any event
}
\end{verbatim}
~
}

@@

As illustrated in Figure~\ref{fig.middleware}, events are broadcast between
peers with a timestamp scheduled to the future with a deadline $\Delta$, such
that all peers are able to apply them in sync.
%
To prevent broadcast cycles, each peer keeps a vector of the highest timestamps
it received from each other peer, such that lower numbers can be ignored when
received.
Since we rely on TCP connections, it is not possible to receive out-of-order
timestamps attached to a given source peer.
When the middleware receives a higher number from a given peer, it updates its
vector position and triggers a broadcast to neighbours.
%
%As mentioned in Section~\ref{sec.tml.api}, we assume that peers assign
%themselves unique identifiers in a contiguous range, such that they fit in a
%simple vector of timestamps.
%Therefore, this is not a severe limitation, and a separate mechanism could be
%used for assignments, with few modifications to the core of the middleware.

@@

{\footnotesize
~
\begin{verbatim}
typedef struct {
    uint8_t  src;       // source peer
    uint32_t tick;      // tick to apply
    p2p_evt  evt;       // event id + payload
} p2p_pak;

struct {
    int i;  // next event to apply in real time
    int n;  // number of events in the queue
    p2p_pak buf[MAX];   // queue of packets
} PAKS;
\end{verbatim}
~
}

@@

{\footnotesize
~
\begin{verbatim}
01  void p2p_travel (int from, int to, int ms) {
02    for (i=from..to) {
03      int bef = <tick-of-snapshot-before-i>
04      <restore-snapshot-at-bef>
05      for (j=bef..i) {
06        cb_sim(<tick-or-evt-at-j>)
07      }
08      cb_ren();
09      <delay-ms>
10    }
11  }
\end{verbatim}
~
}

@@

This section provides an overview of the middleware implementation to answer
how it ensures the desired properties in concrete terms as follows:
    (a) peers meet event deadlines either travelling back and forth or in real
        time (\linx{12--16} or \linx{18--22} / \ref{sec.tml.middleware.loop});
    (b) peers synchronize their clocks with periodic \code{SYNC} packets that
        fast forward late peers
        (\linx{7--10} / \ref{sec.tml.middleware.loop}); and
    (c) peers can join the network at any time and still catch up in time by
        receiving the full event history
        (Section~\ref{sec.tml.middleware.events}).
